R for Clinical Study Reports and Submission

### 2021-08-27 tlf-disposition.Rmd
update output
update content structure
update dispostion table

## Table for Study Patients

In section 10, study patients of a CSR, we need to summarize accounting 
of all patients who entered the study.

@@

The first step is to read relevant datasets into R. 
For disposition table, all the information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

@@

We illustrate how to prepare a report data for a simple disposition of patients table using variables below:

@@

```{r}
adsl %>% select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>% 
         head()
```
```{r}
library(tidyr)
```

```{r}
n_rand <- adsl %>% group_by(TRT01PN) %>%
                   summarize(n = n()) %>% 
                   pivot_wider(names_from = TRT01PN, names_prefix = "n_", values_from = n) %>% 
                   mutate(row = "Patients in population")
n_rand
```

```{r}
n_disc <- adsl %>% group_by(TRT01PN) %>%
                   summarize(n = sum(DISCONFL == "Y"), 
                             pct = formatC(n / n() * 100, digits = 1, format = "f", width = 5)) %>% 
                   pivot_wider(names_from = TRT01PN, values_from = c(n, pct) ) %>% 
                   mutate(row = "Discontinued")
n_disc
```

```{r}
n_reason <- adsl %>% group_by(TRT01PN) %>%
                     mutate(n_total = n()) %>%
                     group_by(TRT01PN, DCREASCD) %>%
                     summarize(n = n(),
                               pct = formatC(n / unique(n_total) * 100, digits = 1, format = "f", width = 5)) %>% 
                     pivot_wider(id_cols = DCREASCD, names_from = TRT01PN, 
                                 values_from = c(n, pct), values_fill = list(n = 0, pct = "  0.0") ) %>% 
                     rename(row = DCREASCD)
n_reason
```

```{r}
n_complete <- n_reason %>% filter(row == "Completed")
n_complete
```

```{r}
n_reason   <- n_reason %>% filter(row != "Completed") %>% 
                  mutate(row = paste0("    ", row))
n_reason
```

@@

```{r}
tbl_disp %>% rtf_body() %>% 
             rtf_encode() %>% 
             write_rtf("tlf/tbl_disp1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp1.pdf")
```

```{r}
adsl %>% select(TRT01P, TRT01PN) %>% 
         unique() %>% 
         arrange(TRT01PN)
```

```{r}
tbl_disp %>% rtf_title("Disposition of Patients")

             rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose", 
                           col_rel_width = c(3, rep(2,3)) ) %>%

             rtf_colheader(" | n | (%) | n | (%) | n | (%)",
                           col_rel_width = c(3, rep(c(0.7, 1.3),3)),
                           border_top = c("", rep("single", 6)),
                           border_left = c("single", rep(c("single",""), 3))) %>%

             rtf_body(col_rel_width      = c(3, rep(c(0.7, 1.3),3)),
                      text_justification = c("l", rep("c",6)),
                      border_left        = c("single", rep(c("single",""), 3))) %>%

             rtf_encode() %>%

             write_rtf("tlf/tbl_disp2.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp2.pdf")
```

>>>>

# Table for Study Patients

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize accounting of all patients who entered the study, 
in the section 10.1, disposition of patients of a CSR.

@@

In this chapter, we illustrate how to create a disposition table as below.

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp.pdf")
```

The first step is to read relevant datasets into R. 
For disposition table, all the required information is saved in the ADSL dataset.
We can use `haven` package to read the dataset.

@@

We illustrate how to prepare a report data for a simplified disposition 
of patients table using variables below:

@@

```{r}
adsl %>%
  select(USUBJID, TRT01P, TRT01PN, DISCONFL, DCREASCD) %>%
  head(4)
```
In the code below, we calculated patients in the analysis population.

```{r}
n_rand <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(n = n()) %>%
  pivot_wider(
    names_from = TRT01PN,
    names_prefix = "n_",
    values_from = n
  ) %>%
  mutate(row = "Patients in population")
n_rand
```
In the code below, we calculate number and percentage of patients who discontinued the study.

Here we use `formatC()` to customize the numeric value to be 1 decimal with fixed width of 5.

```{r}
n_disc <- adsl %>%
  group_by(TRT01PN) %>%
  summarize(
    n = sum(DISCONFL == "Y"),
    pct = formatC(n / n() * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  pivot_wider(
    names_from = TRT01PN,
    values_from = c(n, pct)
  ) %>%
  mutate(row = "Discontinued")

n_disc
```
In the code below, we calculate number and percentage of patients 
who completed/discontinued the study in different reasons.

```{r}
n_reason <- adsl %>%
  group_by(TRT01PN) %>%
  mutate(n_total = n()) %>%
  group_by(TRT01PN, DCREASCD) %>%
  summarize(
    n = n(),
    pct = formatC(n / unique(n_total) * 100,
      digits = 1, format = "f", width = 5
    )
  ) %>%
  pivot_wider(
    id_cols = DCREASCD,
    names_from = TRT01PN,
    values_from = c(n, pct),
    values_fill = list(n = 0, pct = "  0.0")
  ) %>%
  rename(row = DCREASCD)

n_reason
```

In the code below, we calculate number and percentage of patients 
who complete the study. We split `n_reason` because we need to customize 
the row order of the table.

```{r}
n_complete <- n_reason %>% filter(row == "Completed")

n_complete
```
In the code below, we calculate number and percentage of patients 
who discontinued the study in different reasons. 
Here we use `paste0("    ", row)` to add some leading space for individual reasons for study discontinuation.

```{r}
n_reason <- n_reason %>%
  filter(row != "Completed") %>%
  mutate(row = paste0("    ", row))

n_reason
```

Now we combined individual rows into the whole table for reporting purpose. 
`tbl_disp` is used as input for `r2rtf` to create final report.

@@

We start to define the format of the output. We highlighted items that 
is not discussed in previous discussion.

`rtf_title` define table title. We can provide a vector for the title argument. 
Each value is a separate line. 
The format can also be controlled by providing a vector input in text format.

```{r}
tbl_disp %>%

  # Table title
  rtf_title("Disposition of Patients") %>%

  # First row of column header
  rtf_colheader(" | Placebo | Xanomeline Low Dose| Xanomeline High Dose",
    col_rel_width = c(3, rep(2, 3))
  ) %>%

  # Second row of column header
  rtf_colheader(" | n | (%) | n | (%) | n | (%)",
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    border_top = c("", rep("single", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%

  # Table body
  rtf_body(
    col_rel_width = c(3, rep(c(0.7, 1.3), 3)),
    text_justification = c("l", rep("c", 6)),
    border_left = c("single", rep(c("single", ""), 3))
  ) %>%

  # Encoding RTF syntax
  rtf_encode() %>%

  # Save to a file
  write_rtf("tlf/tbl_disp.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("tlf/tbl_disp.pdf")
```

### 2021-08-27 tlf-overview.Rmd
update dispostion table

In the rest of this chapter, we are following the [ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) on 
structure and content of clinical study reports.

@@

```{r}
library(r2rtf)
r2rtf_adae %>%
  select(USUBJID, TRTA, AEDECOD) %>%
  head(4)
```
`dplyr` and `tidyr` packages within `tidyverse` are used 
for data manipulation to create a data frame 
that contains all the information we want to add in an RTF table.

```{r}
library(tidyr)
library(dplyr)
tbl <- r2rtf_adae %>%
  count(TRTA, AEDECOD) %>%
  pivot_wider(names_from = TRTA, values_from = n, values_fill = 0)

@@

```{r}
head(tbl) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%    # define relative width
  rtf_encode() %>%
  write_rtf("tlf/intro-ae2.rtf")
```

@@

```{r}
head(tbl) %>%
  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae3.rtf")
```

@@

```{r}
head(tbl) %>%
  rtf_body(text_justification = c("l", "c", "c", "c")) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae5.rtf")
```

@@

```{r}
head(tbl) %>%
  rtf_colheader(
    colheader = " | Treatment",
    col_rel_width = c(3, 6)
  ) %>%
  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    border_top = c("", "single", "single", "single"),
    col_rel_width = c(3, 2, 2, 2)
  ) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%
  rtf_encode() %>%
  write_rtf("tlf/intro-ae7.rtf")
```

@@

In the r2rtf [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
there are more examples to illustrate how to customize

>>>>

In the rest of this chapter, we are following the 
[ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
on structure and content of clinical study reports.

@@

```{r}
library(r2rtf)

r2rtf_adae %>%
  select(USUBJID, TRTA, AEDECOD) %>%
  head(4)
```
`dplyr` and `tidyr` packages within `tidyverse` are used 
for data manipulation to create a data frame 
that contains all the information we want to add in an RTF table.

```{r}
library(tidyr)
library(dplyr)

tbl <- r2rtf_adae %>%
  count(TRTA, AEDECOD) %>%
  pivot_wider(names_from = TRTA, values_from = n, values_fill = 0)

@@

```{r}
head(tbl) %>%
  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%

  # define relative width
  rtf_encode() %>%
  write_rtf("tlf/intro-ae2.rtf")
```

@@

```{r}
head(tbl) %>%

  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    col_rel_width = c(3, 2, 2, 2)
  ) %>%

  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%

  rtf_encode() %>%

  write_rtf("tlf/intro-ae3.rtf")
```

@@

```{r}
head(tbl) %>%

  rtf_body(text_justification = c("l", "c", "c", "c")) %>%

  rtf_encode() %>%

  write_rtf("tlf/intro-ae5.rtf")
```

@@

```{r}
head(tbl) %>%

  rtf_colheader(
    colheader = " | Treatment",
    col_rel_width = c(3, 6)
  ) %>%

  rtf_colheader(
    colheader = "Adverse Events | Placebo | Xanomeline High Dose | Xanomeline Low Dose",
    border_top = c("", "single", "single", "single"),
    col_rel_width = c(3, 2, 2, 2)
  ) %>%

  rtf_body(col_rel_width = c(3, 2, 2, 2)) %>%

  rtf_encode() %>%

  write_rtf("tlf/intro-ae7.rtf")
```

@@

In the `r2rtf` R package [get started](https://merck.github.io/r2rtf/articles/r2rtf.html) page
there are more examples to illustrate how to customize
